---
title: "Collate RNA counts and quantify proportion in fractions"
author: "Edward Wallace"
date: "26th September 2017"
output:
  html_document:
    toc: true
    toc_depth: 4
---


This scripts takes as inputs read counts on ORFs from TSP datasets.
    
Outputs for select (verified) ORFs:
    - collected read counts, in integer counts.
    - pSup estimates for proportion in supernatant, a fraction between 0 and 1.


```{r setup,warning=FALSE,message=FALSE,echo=FALSE}

## knitr options for report generation
knitr::opts_chunk$set(warning=FALSE,message=FALSE,echo=TRUE,cache=FALSE,
                      results="show",
                      fig.path="figure/quantFromCounts-",
                      cache.path="cache/quantFromCounts-")


## data processing options, packages, functions
options(stringsAsFactors=FALSE)  # load character strings as strings
library(plyr)
library(reshape2) # data manipulation
library(tidyverse) # more data manipulation
library(stringr)


library(rstan) # Bayesian model description and fitting
# run in parallel
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

library(cowplot) # extensions and nice defaults
 # set default theme for graphics
# theme_set(theme_bw() %+replace% theme(panel.grid.minor=element_blank())) 
theme_set(theme_cowplot(font_size=11) %+replace% 
              theme(# panel.grid.major=element_line(size=0.15,colour="grey50"),
                  panel.border=element_rect(colour = "grey50",
                                            linetype = "solid",size=0.5),
                    strip.background = element_blank()))


geom_diagline <- function(linetype='solid',size=0.1,colour="grey20",...) {
    geom_abline(slope=1,intercept=0,linetype=linetype,colour=colour)
}

scientific_10 <- function(x) {
    xout <- gsub("1e", "10^{", format(x),fixed=TRUE)
    xout <- gsub("{-0", "{-", xout,fixed=TRUE)
    xout <- gsub("{+", "{", xout,fixed=TRUE)
    xout <- gsub("{0", "{", xout,fixed=TRUE)
    xout <- paste(xout,"}",sep="")
    return(parse(text=xout))
}

scale_x_log10nice <- function(name=waiver(),omag=seq(-10,20),...) {
    breaks10 <- 10^omag
    scale_x_log10(name,breaks=breaks10,labels=scientific_10(breaks10),...)
}

scale_y_log10nice <- function(name=waiver(),omag=seq(-10,20),...) {
    breaks10 <- 10^omag
    scale_y_log10(name,breaks=breaks10,labels=scientific_10(breaks10),...)
}

scale_loglog <- function(...) {
    list(scale_x_log10nice(...),scale_y_log10nice(...))
}

scale_x_log2nice <- function(name=waiver(),omag=seq(-6,6),...) {
    breaks2 <- 2^omag
    scale_x_log10(name,breaks=breaks2,
                  labels=parse(text=paste("2^{",omag,"}")),...)
}

scale_y_log2nice <- function(name=waiver(),omag=seq(-6,6),...) {
    breaks2 <- 2^omag
    scale_y_log10(name,breaks=breaks2,
                  labels=parse(text=paste("2^{",omag,"}")),...)
}

scale_loglog2 <- function(...) {
    list(scale_x_log2nice(...),scale_y_log2nice(...))
}

theme_density <- function(...) {
    list(scale_y_continuous(expand=c(0.01,0.01)),
         theme(panel.border=element_blank(),
               panel.background=element_blank(),
               axis.line=element_blank(),
               axis.text.y=element_blank(),
               axis.title.y=element_blank(),
               axis.ticks.y=element_blank()),...)
}

corfinite <- function(x,y,method="pearson") {
    # wrapper to calculate correlation coefficients for only finite values
    # useful for correlations of log-transformed values
    niceinds <- which(is.finite(x) & is.finite(y))
    cor(x[niceinds],y[niceinds],method=method)
}

logcor <- function(x,y,method="pearson") {
    # wrapper for correlations of log-transformed values
    corfinite(log(x),log(y),method=method)
}

txtheader <- 
"# HeatShock Quinary mRNA
# Edward Wallace September 2017, Edward.Wallace@ed.ac.uk
# 
"
```


# Total, Supernatant, and 100,000g Pellet (TSP) RNA data

## Load data

```{r load_data}
# setwd("~/Repos/RNAFracQuant/script_initial/")
Samplesheet <- read_tsv("../data_in/TSP_count/Samplesheet.txt",comment="#")
ORFselect <- read_tsv("../data_in/orf_coding_EW_select.txt")
dir_processed <- "../data_processed/TSP_count"

read_ORFCount <- function(file_ext,dir_in) {
    # Convenience function for reading data in given format
    paste0(dir_in,"/",file_ext) %>% 
        read_tsv(
            col_names = c("ORF","Count"),
            comment="__")
}

# Load all data from all the files, with metadata
TSP_Count <- Samplesheet %>%
    group_by(File,Sample,Condition,Fraction) %>%
    do(read_ORFCount(file_ext=.$File[1],dir_in="../data_in/TSP_count/")) %>%
    mutate(ORF=stringr::str_sub(ORF,end=-5))

# Filter for only selected ORFs, and calculate TPMs
TSP_Count_TPM <- TSP_Count %>%
    inner_join(ORFselect) %>%
    group_by(Sample,Condition,Fraction) %>%
    mutate(Density=Count/Length, 
           TPM=Density/sum(Density)*1e6  ) %>%
    select(Sample,Condition,Fraction,ORF,Count,TPM)

TSP_Count_TPM
# # check TPM sums to 1mi.
TSP_Count_TPM %>%
    summarize(Count=sum(Count),TPM=sum(TPM))

cat(txtheader,
"# TSP_Count_TPM_tidy.txt
# TSP measurements, RNA Counts and TPM
# Tidy format
# 
",
    file=paste0(dir_processed,"TSP_Count_TPM_tidy.txt"),sep="")
write_tsv(TSP_Count_TPM %>%
              mutate(TPM=round(TPM,1)),
          paste0(dir_processed,"TSP_Count_TPM_tidy.txt"),
          append=TRUE,col_names=TRUE)
```

### Make wide TSP

```{r TPM_wide,dependson="load_data"}
TSP_TPM_halfwide <- 
    TSP_Count_TPM %>%
    ungroup() %>%
    select(Condition,Fraction,ORF,TPM) %>%
    spread(Fraction,TPM)

TSP_TPM_halfwide

cat(txtheader,
"# TSP_TPM_halfwide.txt
# TSP measurements, TPM
# Half-wide, collected by ORF.
# One column per fraction, collected by Condition
# 
",
    file=paste0(dir_processed,"TSP_TPM_halfwide.txt"),sep="")
write_tsv(TSP_Count_TPM %>%
              ungroup() %>%
              select(Condition,Fraction,ORF,TPM) %>%
              mutate(TPM=round(TPM,1)) %>%
              spread(Fraction,TPM),
          paste0(dir_processed,"TSP_TPM_halfwide.txt"),
          append=TRUE,col_names=TRUE)

TSP_TPM_wide <- 
    TSP_Count_TPM %>%
    ungroup() %>%
    select(Condition,Fraction,ORF,TPM) %>%
    unite(Var,Condition,Fraction,sep="_") %>%
    spread(Var,TPM)

TSP_TPM_wide

cat(txtheader,
"# TSP_TPM_wide.txt
# TSP measurements, TPM
# Wide, collected by ORF.
# Each column is TPM for that ORF in a given Condition and Fraction
# 
",
    file=paste0(dir_processed,"TSP_TPM_wide.txt"),sep="")
write_tsv(TSP_Count_TPM %>%
              ungroup() %>%
              select(Condition,Fraction,ORF,TPM) %>%
              unite(Var,Condition,Fraction,sep="_") %>%
              mutate(TPM=round(TPM,1)) %>%
              spread(Var,TPM),
          paste0(dir_processed,"TSP_TPM_wide.txt"),
          append=TRUE,col_names=TRUE)
```

## Estimate proportion in 100,000g Supernatant (pSup) for TSP


#### STAN functions to fit mixing ratios

```{r mixingratiosTSP_stan}

mdat_mixingTSP <- function(ctdata) {
    ## make data for stan fit
    head(ctdata)
    list(NRNA=nrow(ctdata),
         tot_obs=as.integer(round(ctdata$Tot)),
         sup_obs=as.integer(round(ctdata$Sup)),
         p100_obs=as.integer(round(ctdata$P100))
         )
    
}

make_mixingTSP <- function(ctdata) {
    stan_dat <- mdat_mixingTSP(ctdata)
    stan(model_code='// -*- mode: C -*-
data {
  // Number of RNAs
  int<lower=1> NRNA;     
  
  // Note: These are all integers
  // columns t, s, p
  int<lower=0> tot_obs[NRNA];
  int<lower=0> sup_obs[NRNA];
  int<lower=0> p100_obs[NRNA];
}
parameters {
  // Unnormalized mixing proportions
  // real<lower=0> mixing_t;
  real<lower=0> mixing_sup;
  real<lower=0> mixing_p100;
  
  // dispersion parameter for counts
  real phi;
}
model{
  // mixing ratios
  mixing_sup ~ gamma(1,1);
  mixing_p100 ~ gamma(1,1);
  // Cauchy prior for negbin dispersion parameter
  phi ~ cauchy(0,3);
  
  for(idx in 1:NRNA){ 
    // count distn negative binomial with specified means
    // Total
    tot_obs[idx] ~ neg_binomial_2(mixing_sup * sup_obs[idx] + mixing_p100 * p100_obs[idx], phi);
  }

}
generated quantities{
  // print("Mixing pars (sup,p100) = (", mixing_sup,",",mixing_p100,")");
  // print("dispersion phi = ", phi);
  // print("------------------");
}
',
data=stan_dat,chains = 1,iter = 10)
}

fit_mixingTSP <- function(ctdata,stan_mixing=NULL,...) {
    stan_dat <- mdat_mixingTSP(ctdata)
    if (is.null(stan_mixing)) {
        stan_mixing <- make_mixingTSP(ctdata)
    }
    stan_mixing_fit <- stan(fit=stan_mixing,data=stan_dat,chains = 4,...)
    return(stan_mixing_fit)
}

getmixingratiosTSP <- function(ctdata,iter=1000,
                                control=list(adapt_delta=0.85),...) {
    # head(ctdata) %>% print()
    stansummary <- fit_mixingTSP(ctdata=ctdata,iter=iter,control=control,...) %>%
        summary()
    # return medians
    data.frame(
        mixing.Sup=stansummary$summary["mixing_sup","50%"],
        mixing.P100=stansummary$summary["mixing_p100","50%"],
        lp.n_eff  =stansummary$summary["lp__","n_eff"],
        lp.Rhat   =stansummary$summary["lp__","Rhat"])
}

```

#### Fit mixing ratios from well detected mRNA 

Filter by median tpm > 5, min tpm > 0, across all conditions.

This is to remove low-abundance transcripts, where the counts aren't informative anyway, and reduce the time to fit.

```{r fitmixingratiosTSP,dependson=c("TSP_wide","mixingratiosTSP_stan")}

wd5_TSP_Tot <- 
    TSP_Count_TPM %>%
    filter(Fraction=="Tot") %>%
    group_by(ORF) %>%
    summarise(d5 = sum(TPM > 5),wd5=(d5==4)) %>%
    filter(wd5) %>%
    .$ORF

TSP_Count_FracBySample <- 
    TSP_Count_TPM %>%
    ungroup() %>%
    select(Condition,Fraction,ORF,Count) %>%
    spread(Fraction,Count) %>%
    filter(ORF %in% wd5_TSP_Tot)

# compile stan model, test it with 10 iterations, save output
stan_mixingTSP_wd5 <- TSP_Count_FracBySample %>%
    make_mixingTSP()

# Specify a seed for random number generator so output is reproducible
myseed = 39

# run stan mixing ratio inference, reusing same model
mixing_ratios_TSP <- TSP_Count_FracBySample %>%
    ddply(~Condition,getmixingratiosTSP,
          stan_mixing=stan_mixingTSP_wd5,iter=1000,seed=myseed)

mixing_ratios_TSP

cat(txtheader,
"# mixing_ratios_TSP100.txt
# mixing ratios and convergence estimate for TSP100
# 
",
    file=paste0(dir_processed,"mixing_ratios_TSP100.txt"),sep="")
write_tsv(mixing_ratios_TSP,
          paste0(dir_processed,"mixing_ratios_TSP100.txt"))
```

#### Calculate pSup, plot *counts* to compare

```{r calcpSupetcTSP,dependson="fitmixingratiosTSP",fig.height=4.5,fig.width=8}
TSP_Count_FracBySample_pSup <- 
    TSP_Count_FracBySample %>%
    left_join(select(mixing_ratios_TSP,Condition,mixing.Sup,mixing.P100)) %>%
    mutate(mixSum=mixing.Sup*Sup+mixing.P100*P100,
           pSup=mixing.Sup*Sup/mixSum)


TSP_Count_FracBySample_pSupcor <- 
    TSP_Count_FracBySample_pSup %>%
    group_by(Condition) %>%
    summarise(R.mixSum=logcor(Tot,mixSum),
              R.Sup=logcor(Tot,Sup),
              R.P100=logcor(Tot,P100),
              R.SupP100=logcor(Sup,P100))


cat(txtheader,
"# TSP_Count_FracBySample_pSupcor.txt
# Correlation between total and fractions, TSP
# 
",
    file=paste0(dir_processed,"TSP_Count_FracBySample_pSupcor.txt"),sep="")
write_tsv(TSP_Count_FracBySample_pSupcor %>%
              mutate(R.mixSum=round(R.mixSum,4),
                     R.Sup=round(R.Sup,4),
                     R.P100=round(R.P100,4),
                     R.SupP100=round(R.SupP100,4)) ,
          paste0(dir_processed,"TSP_Count_FracBySample_pSupcor.txt"),
          append=TRUE,col_names=TRUE)

```

```{r writepSupetcTSP,dependson=c("calcpSupetcTSP","TSP_wide"),fig.height=4.5,fig.width=8}
TSP_Count_FracBySample_pSup_adj <- 
    TSP_Count_FracBySample_pSup %>%
    mutate(TPM=round(pSup,4)) %>% # This is a convenient lie; pSup is not a TPM
    mutate(Fraction="pSup") %>%
    select(Condition,Fraction,ORF,TPM) 
    

cat(txtheader,
"# TSP_TPM_pSup_wd5_halfwide.txt
# TSP measurements of RNA TPM and proportion in 100,000g supernatant
# Restricted to well-detected ORFS, > 5 TPM in all Tot fractions
# 
# Half-wide, collected by ORF.
# One column per measurement/estimate, collected by Condition and Biorep
# Tot, Sup, P100 are RNA abundances in Transcripts per Million (TPM)
# pSup/proportion in Supernatate is a*Sup / (a*Sup+b*P100), where weights
#   a & b are estimated by a best-fit of Tot = a*Sup + b*P100
# 
",
    file=paste0(dir_processed,"TSP_TPM_pSup_wd5_halfwide.txt"),sep="")
write_tsv(TSP_Count_TPM %>%
              ungroup() %>%
              filter(ORF %in% wd5_TSP_Tot) %>%
              select(Condition,Fraction,ORF,TPM) %>%
              mutate(TPM=round(TPM,1),Fraction=as.character(Fraction)) %>%
              bind_rows(TSP_Count_FracBySample_pSup_adj) %>%
              mutate(Fraction=factor(Fraction, 
                                     levels=c("Tot","Sup","P100","pSup"))
                     ) %>%
              spread(Fraction,TPM),
          paste0(dir_processed,"TSP_TPM_pSup_wd5_halfwide.txt"),
          append=TRUE,col_names=TRUE)

# Write out all data

cat(txtheader,
"# TSP_TPM_pSup_halfwide.txt
# TSP measurements of RNA TPM and proportion in 100,000g supernatant
# 
# Half-wide, collected by ORF.
# One column per measurement/estimate, collected by Condition and Biorep
# Tot, Sup, P100 are RNA abundances in Transcripts per Million (TPM)
# pSup/proportion in Supernatate is a*Sup / (a*Sup+b*P100), where weights
#   a & b are estimated by a best-fit of Tot = a*Sup + b*P100
# 
",
    file=paste0(dir_processed,"TSP_TPM_pSup_halfwide.txt"),sep="")
write_tsv(TSP_Count_TPM %>%
              ungroup() %>%
              select(Condition,Fraction,ORF,TPM) %>%
              mutate(TPM=round(TPM,1),Fraction=as.character(Fraction)) %>%
              bind_rows(TSP_Count_FracBySample_pSup_adj) %>%
              mutate(Fraction=factor(Fraction, 
                                     levels=c("Tot","Sup","P100","pSup"))
                     ) %>%
              spread(Fraction,TPM),
          paste0(dir_processed,"TSP_TPM_pSup_halfwide.txt"),
          append=TRUE,col_names=TRUE)
```

```{r plotpSupetcTSP,dependson="calcpSupetcTSP",fig.height=2.5,fig.width=7.5}

ggplot(data=TSP_Count_FracBySample_pSup,aes(x=Tot,y=Sup)) +
    geom_diagline() + 
    geom_point(size=1) + 
    scale_loglog() + 
    geom_text(data=TSP_Count_FracBySample_pSupcor,
              aes(label=paste("R =",round(R.Sup,3)),x=8e2,y=3e5)) +
    facet_wrap(~Condition,ncol=4)

ggplot(data=TSP_Count_FracBySample_pSup,aes(x=Tot,y=P100)) +
    geom_diagline() + 
    geom_point(size=1) + 
    scale_loglog() + 
    geom_text(data=TSP_Count_FracBySample_pSupcor,
              aes(label=paste("R =",round(R.P100,3)),x=8e2,y=3e5)) +
    facet_wrap(~Condition,ncol=4)


ggplot(data=TSP_Count_FracBySample_pSup,aes(x=Tot,y=mixSum)) +
    geom_diagline() + 
    geom_point(size=1) + 
    scale_loglog() + 
    geom_text(data=TSP_Count_FracBySample_pSupcor,
              aes(label=paste("R =",round(R.mixSum,3)),x=8e2,y=3e5)) +
    facet_wrap(~Condition,ncol=4)
```


The mixSum correlation shows that the re-fitting works very well.

